---
title: "Terrain Analyses in R"
author: "Matthew Ross"
date: "2024-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(sf)
library(terra)
library(whitebox)
library(elevatr)
library(mapview)
library(tidyverse)
library(tmap)


# Run first time you install whitebox
# whitebox::install_whitebox
```


# Goals


## Terrain Analysis Toolkit

The goal of this lab is to familiarize you with a very powerful set of tools
that enable you to do terrain analyses. These are primarily the packages `elevatr` for
downloading DEM data, `whitebox` for conducting watershed and terrain analyses.
These packages also rely on other geospatial packages (`sf`, `terra`) and visualizing
the data relies on `mapview` and `tmap`. 


## Terrain Analysis Ideas

So the above tools can help you analyze a digital elevation model, but why would
you want to do this? Well, terrain analysis can do many things in water resource
planning and analysis. For example, if we wanted to find parts of the landscape
where we think wetlands could exist, we might want to use [Topographic Wetness 
Index](https://www.whiteboxgeo.com/manual/wbt_book/available_tools/geomorphometric_analysis.html#wetnessindex) 
to identify these areas where the topography lends itself to wet landscapes, which
are areas of relatively high flow accumulation coupled with a bowl-like curvature.

To learn these tools and ideas we will work through a watershed delineation example
in the South Fork of the Poudre River, and then you will conduct your own analysis. 


# South Fork Example

## Add a watershed outlet point

```{r}

#Create a point near the stream using lat, long. 
sheds <- tibble(site = c('South Fork'),
                long = c(40.5475185),
                lat = c(-105.6091385)) %>%
  st_as_sf(., coords = c('lat','long'), crs = 4263) %>%
  st_transform(2163)

#Visualize the points
mapview(sheds)

#Little code snippet to make the data folder if you don't have it. 
if(!file.exists('data')){
  dir.create('data')
}

st_write(sheds, 'data/pourpoints.shp', append = F)
```


## Get elevation data from around that area. 

```{r}

# Use elevatr to download data and convert to terra object
elev <- get_elev_raster(sheds, z = 10) %>%
  rast(.)

# Various ways to plot

mapview(elev) + 
  mapview(sheds)
```

### Save a local DEM Version

```{r}
#Save local version for whitebox to use
writeRaster(elev, 'data/elev.tif', overwrite = T)
```



## Get hillshade for display purposes

```{r}

#Get a shillshade map from wbt
wbt_hillshade(dem = "data/elev.tif",
              output = "data/hillshade.tif",
              azimuth = 115)

#Read in the hillshade
hillshade <- rast('data/hillshade.tif')

#Visualize
tm_shape(hillshade)+
  tm_raster(style = "cont",
            palette = "-Greys", 
            legend.show = FALSE)+
  tm_scale_bar()

mapview(hillshade) +
  mapview(sheds)
```


## Hydrologically condition DEM

JP Gannon does a great job on his [hydroinformatics website](https://vt-hydroinformatics.github.io/Quarto_Book/14-Geospatial-Raster-Hydro.html) explaining why we need to hydrologically condition a DEM before we can conduct
watershed analyses. Basically, DEMs aren't perfect and they can have artifacts 
in them that make it so that water doesn't properly flow downhill. To force water
to move the way we think it should through the landscape we can "condition" these
DEMs using the functions below. 

```{r}

# Breach depressions (force water to move through ponds/lakes/etc...)
wbt_breach_depressions_least_cost(
  dem = "data/elev.tif",
  output = "data/breached.tif",
  dist = 9,
  fill = TRUE)


# Fill any remaining depressions
wbt_fill_depressions_wang_and_liu(
  dem = "data/breached.tif",
  output = "data/breachfill.tif")
```

## Flow Accumulation 

Flow accumulation is the key function we use to estimate how much watershed area
is draining to a specific point anywhere in the river network. Here we use the D8
flow algorithm, which makes it so 100% of the water is routed to nearby cells. 
ESRI has a nice article visualizing what this algorithm is [doing](https://pro.arcgis.com/en/pro-app/latest/tool-reference/raster-analysis/flow-direction.htm)

```{r}

#Get flow accumulation
wbt_d8_flow_accumulation(input = "data/breachfill.tif",
                         output = "data/d8fa.tif")
#Get flow direction
wbt_d8_pointer(dem ='data/breachfill.tif',
               output = 'data/d8point.tif')

# read in the data
fa <- rast('data/d8fa.tif') %>%
  log10(.)

#visualize
mapview(fa) + 
  mapview(sheds)
```


## Extract streams with arbitrary 300 cell threshold 

In order to make sure that our watershed outlet point, which we arbitrarily and
manually extracted lats and longs for, we need to first extract a stream network.
We are picking a 300-cell flow accumulation threshold for our 'stream initiation.' 
Because our starting raster is a ~58X58m cell size 300 cells roughly equals 1 km2.
This means, we are saying that once a stream has more than 1 km2 of watershed area draining
to that point in the stream, we think that it will have a functioning stream channel.
You can conduct entire studies to get this threshold right, so this is arbitrary here. 

```{r}

#Extract streams at 1km2 
wbt_extract_streams(flow_accum = "data/d8fa.tif",
                    output = "data/raster_streams.tif",
                    threshold = 300)

#Snap our watershed flowliens to this point. 
wbt_jenson_snap_pour_points(pour_pts = "data/pourpoints.shp",
                            streams = "data/raster_streams.tif",
                            output = "data/snappedpp.shp",
                            snap_dist = 200)

# Read in and check the snap. 
snap_pour <- st_read('data/snappedpp.shp')
mapview(fa) + 
  mapview(snap_pour)

```


## Delineate watershed


Finally after all that! We can delineate our watershed


```{r}
wbt_watershed(d8_pntr = "data/d8point.tif",
              pour_pts = "data/snappedpp.shp",
              output = "data/south_fork.tif")

south_shed <- rast('data/south_fork.tif')

mapview(south_shed) + 
mapview(sheds)

```

## Extract a terrain metric about this watershed

### Generate terrain metric

Delineating a watershed allows us to now get watershed terrain metrics for that 
watershed. Whitebox tools has dozens of terrain tools you can use
[See More in the Geomorph Section](https://www.whiteboxgeo.com/manual/wbt_book/intro.html).
Here we will generate and extract one critical variable. Topographic Wetness,
which indicates areas in a watershed (or a watershed average) wetness. High values
indicate areas that can be marshy/wetlandy and generally wet. Low values (like ridge tops) will
be drier parts of the landscape. 

```{r}
#First we need slope
wbt_slope(dem = 'data/elev.tif',
          output = 'data/slope.tif',
          units = 'degrees')

# Now we can generate twi
wbt_wetness_index(sca = 'data/d8fa.tif',
                  slope = 'data/slope.tif',
                  output = 'data/twi.tif')

twi <- rast('data/twi.tif')

mapview(twi)


```

### Crop to watershed and extract average value


```{r}
# Crop TWI to south_shed area

twi_south <- crop(twi, south_shed) 


twi_south_mask <- mask(twi, south_shed)


mapview(twi_south_mask)


global(twi_south_mask, fun = 'mean', na.rm = T)
```



# Assignment


## Q1 Generate your own watershed

Using the above code as a starter, generate your own watershed. Try to only
pick a watershed that is not too large in size (~ < 1000km2). Remember
you may need to mess around with the Z level for get_elev_raster(). 


```{r}

#Create a point near the stream using lat, long. 
fr_sheds <- tibble(site = c("Le Giffre",
                            "Le Bout du Monde"),
                lat_long = c("46.0566073, 6.7806049",
                             "46.1088149, 6.8626079")) %>%
  separate(lat_long, into = c("lat", "long"), sep = ",") %>% 
  st_as_sf(., coords = c("long", "lat"), crs = 4263) %>% # crs is still 4263
  st_transform(2154) # search epsg code for France

mapview(fr_sheds) # check map for point locations

#Little code snippet to make the data folder if you don't have it. 
if(!file.exists('data')){
  dir.create('data')
}

# creating space to save shapefile that will be generated later
st_write(fr_sheds, 'data/fr_pourpoints.shp', append = F) 

# download raster from open data (using elevatr package - see CRAN for more information)
fr_elev <- get_elev_raster(fr_sheds, z = 12) %>% 
  rast()
# type fr_elev into console to see data for the object
# plot cool zoomable map of our raster and the points on it

mapview(fr_elev) +
  mapview(fr_sheds)

# save as a DEM locally - place raster in data folder, which must be in gitignore
# from Terra package
writeRaster(fr_elev, "data/fr_elev.tif",
            overwrite = TRUE) 
# added overwrite = true bc I had to keep re-running this code and the error was stopping it

# add hillshade - in hydroinformatics textbook, find hillshade example in watershed delineation chapter
wbt_hillshade(dem = "data/fr_elev.tif",
              output = "data/fr_hillshade.tif",
              azimuth = 115)

#Read in the hillshade
fr_hillshade <- rast('data/fr_hillshade.tif')

#Visualize
tm_shape(fr_hillshade) +
  tm_raster(style = "cont",
            palette = "-Greys", 
            legend.show = FALSE) +
  tm_scale_bar()

mapview(fr_hillshade) +
  mapview(fr_sheds)


# prep DEM for watershed delineation

# white box tools to breach any local depressions to create a stream network
wbt_breach_depressions_least_cost(
  dem = "data/fr_elev.tif",
  output = "data/fr_breached.tif",
  dist = 9,
  fill = TRUE)


# Fill any remaining depressions
wbt_fill_depressions_wang_and_liu(
  dem = "data/fr_breached.tif",
  output = "data/fr_breachedfill.tif")

# Flow accumulation tools
wbt_d8_flow_accumulation(input = "data/fr_breachedfill.tif",
                         output = "data/fr_d8fa.tif")

# Get flow direction - D8 method sends flow to most downslope neighboring square 
# be there are 8 surrounding squares
wbt_d8_pointer(dem = "data/fr_breachedfill.tif",
               output = "data/fr_d8point.tif")

# read in the data
fr_fa <- rast("data/fr_d8fa.tif") %>%
  log10()

#visualize
mapview(fr_fa) + 
  mapview(fr_sheds)

# Extract streams at 1km2 
# my raster has a resolution of 13x13m, so 1km2 is ~ 6000 cells!
wbt_extract_streams(flow_accum = "data/fr_d8fa.tif",
                    output = "data/raster_frstreams.tif",
                    threshold = 6000) # this is the number of cells

#Snap our watershed flowlines to this point. 
wbt_jenson_snap_pour_points(pour_pts = "data/fr_pourpoints.shp",
                            streams = "data/raster_frstreams.tif",
                            output = "data/fr_snappedpp.shp",
                            snap_dist = 200)

# Read in and check the snap. 
fr_snap_pour <- st_read("data/fr_snappedpp.shp")

mapview(fr_fa) + 
  mapview(fr_snap_pour)

# watershed delienation time!
wbt_watershed(d8_pntr = "data/fr_d8point.tif",
              pour_pts = "data/fr_snappedpp.shp",
              output = "data/griffe.tif")

griffe_shed <- rast("data/griffe.tif")

mapview(griffe_shed) +
  mapview(fr_sheds)
```


## Generate, visualize, and extract two additional terrain metrics (not TWI)

```{r}
# Find ridges tool
# I wanted to see how this tool works bc at my job I had to manually 
# mark ridges in 2D model domains as breaklines and I would love if it just worked

wbt_find_ridges(dem = "data/fr_elev.tif",
                output = "data/fr_ridges.tif")

ridges <- rast("data/fr_ridges.tif")

mapview(fr_hillshade) +
  mapview(ridges) +
  mapview(fr_sheds)

# this tool did not work as well as I wish it did, but is a good start.
```



```{r}
# Terrain ruggedness index - 
# this factor was not very interesting, but straightforward to generate

wbt_ruggedness_index(dem = "data/fr_elev.tif",
                     output = "data/fr_ruggedness.tif")

ruggedness <- rast("data/fr_ruggedness.tif")

mapview(ruggedness) +
  mapview(fr_sheds)
```



I have not been able to make the following code work, I'm getting an error saying it panicked
```{r}
# Sediment transport index! I chose this because sediment transfer is not considered
# in effective FEMA floodplain mapping, like i did at work
# so I never learned to model sediment transfer

# this function requires inputs of two other rasters:
# sca - specific contributing area file, which is actually just flow accumulation raster
# slope raster file

wbt_slope(dem = 'data/fr_elev.tif',
          output = 'data/fr_slope.tif',
          units = 'degrees')

wbt_sediment_transport_index(sca = "data/fr_fa.tif",
                             slope = "data/fr_slope.tif",
                             output = "data/fr_sediment.tif")

sed <- rast("data/fr_sediment.tif")

mapview(sed)
```

