---
title: "Terrain Analyses in R"
author: "Matthew Ross"
date: "2024-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(sf)
library(terra)
library(whitebox)
library(elevatr)
library(mapview)
library(tidyverse)
library(tmap)


# Run first time you install whitebox
# whitebox::install_whitebox
```

# Goals

## Terrain Analysis Toolkit

The goal of this lab is to familiarize you with a very powerful set of tools that enable you to do terrain analyses. These are primarily the packages `elevatr` for downloading DEM data, `whitebox` for conducting watershed and terrain analyses. These packages also rely on other geospatial packages (`sf`, `terra`) and visualizing the data relies on `mapview` and `tmap`.

## Terrain Analysis Ideas

So the above tools can help you analyze a digital elevation model, but why would you want to do this? Well, terrain analysis can do many things in water resource planning and analysis. For example, if we wanted to find parts of the landscape where we think wetlands could exist, we might want to use [Topographic Wetness Index](https://www.whiteboxgeo.com/manual/wbt_book/available_tools/geomorphometric_analysis.html#wetnessindex) to identify these areas where the topography lends itself to wet landscapes, which are areas of relatively high flow accumulation coupled with a bowl-like curvature.

To learn these tools and ideas we will work through a watershed delineation example in the South Fork of the Poudre River, and then you will conduct your own analysis.

# South Fork Example

## Add a watershed outlet point

```{r}

#Create a point near the stream using lat, long. 
sheds <- tibble(site = c('South Fork'),
                long = c(40.5475185),
                lat = c(-105.6091385)) %>%
  st_as_sf(., coords = c('lat','long'), crs = 4263) %>%
  st_transform(2163)

#Visualize the points
mapview(sheds)

#Little code snippet to make the data folder if you don't have it. 
if(!file.exists('data')){
  dir.create('data')
}

st_write(sheds, 'data/pourpoints.shp', append = F)
```

## Get elevation data from around that area.

```{r}

# Use elevatr to download data and convert to terra object
elev <- get_elev_raster(sheds, z = 10) %>%
  rast(.)

# Various ways to plot

mapview(elev) + 
  mapview(sheds)
```

### Save a local DEM Version

```{r}
#Save local version for whitebox to use
writeRaster(elev, 'data/elev.tif', overwrite = T)
```

## Get hillshade for display purposes

```{r}

#Get a shillshade map from wbt
wbt_hillshade(dem = "data/elev.tif",
              output = "data/hillshade.tif",
              azimuth = 115)

#Read in the hillshade
hillshade <- rast('data/hillshade.tif')

#Visualize
tm_shape(hillshade)+
  tm_raster(style = "cont",
            palette = "-Greys", 
            legend.show = FALSE)+
  tm_scale_bar()
```

## Hydrologically condition DEM

JP Gannon does a great job on his [hydroinformatics website](https://vt-hydroinformatics.github.io/Quarto_Book/14-Geospatial-Raster-Hydro.html) explaining why we need to hydrologically condition a DEM before we can conduct watershed analyses. Basically, DEMs aren't perfect and they can have artifacts in them that make it so that water doesn't properly flow downhill. To force water to move the way we think it should through the landscape we can "condition" these DEMs using the functions below.

```{r}

# Breach depressions (force water to move through ponds/lakes/etc...)
wbt_breach_depressions_least_cost(
  dem = "data/elev.tif",
  output = "data/breached.tif",
  dist = 9,
  fill = TRUE)


# Fill any remaining depressions
wbt_fill_depressions_wang_and_liu(
  dem = "data/breached.tif",
  output = "data/breachfill.tif"
)
```

## Flow Accumulation

Flow accumulation is the key function we use to estimate how much watershed area is draining to a specific point anywhere in the river network. Here we use the D8 flow algorithm, which makes it so 100% of the water is routed to nearby cells. ESRI has a nice article visualizing what this algorithm is [doing](https://pro.arcgis.com/en/pro-app/latest/tool-reference/raster-analysis/flow-direction.htm)

```{r}

#Get flow accumulation
wbt_d8_flow_accumulation(input = "data/breachfill.tif",
                         output = "data/d8fa.tif")
#Get flow direction
wbt_d8_pointer(dem ='data/breachfill.tif',
               output = 'data/d8point.tif')

# read in the data
fa <- rast('data/d8fa.tif') %>%
  log10(.)

#visualize
mapview(fa) + 
  mapview(sheds)
```

## Extract streams with arbitrary 300 cell threshold

In order to make sure that our watershed outlet point, which we arbitrarily and manually extracted lats and longs for, we need to first extract a stream network. We are picking a 300-cell flow accumulation threshold for our 'stream initiation.' Because our starting raster is a \~58X58m cell size 300 cells roughly equals 1 km2. This means, we are saying that once a stream has more than 1 km2 of watershed area draining to that point in the stream, we think that it will have a functioning stream channel. You can conduct entire studies to get this threshold right, so this is arbitrary here.

```{r}

#Extract streams at 1km2 
wbt_extract_streams(flow_accum = "data/d8fa.tif",
                    output = "data/raster_streams.tif",
                    threshold = 300)

#Snap our watershed flowliens to this point. 
wbt_jenson_snap_pour_points(pour_pts = "data/pourpoints.shp",
                            streams = "data/raster_streams.tif",
                            output = "data/snappedpp.shp",
                            snap_dist = 200)

# Read in and check the snap. 
snap_pour <- st_read('data/snappedpp.shp')
mapview(fa) + 
  mapview(snap_pour)

```

## Delineate watershed

Finally after all that! We can delineate our watershed

```{r}
wbt_watershed(d8_pntr = "data/d8point.tif",
              pour_pts = "data/snappedpp.shp",
              output = "data/south_fork.tif")

south_shed <- rast('data/south_fork.tif')

mapview(south_shed) + 
mapview(sheds)

```

## Extract a terrain metric about this watershed

### Generate terrain metric

Delineating a watershed allows us to now get watershed terrain metrics for that watershed. Whitebox tools has dozens of terrain tools you can use [See More in the Geomorph Section](https://www.whiteboxgeo.com/manual/wbt_book/intro.html). Here we will generate and extract one critical variable. Topographic Wetness, which indicates areas in a watershed (or a watershed average) wetness. High values indicate areas that can be marshy/wetlandy and generally wet. Low values (like ridge tops) will be drier parts of the landscape.

```{r}
#First we need slope
wbt_slope(dem = 'data/elev.tif',
          output = 'data/slope.tif',
          units = 'degrees')

# Now we can generate twi
wbt_wetness_index(sca = 'data/d8fa.tif',
                  slope = 'data/slope.tif',
                  output = 'data/twi.tif')

twi <- rast('data/twi.tif')

mapview(twi)


```

### Crop to watershed and extract average value

```{r}
# Crop TWI to south_shed area

twi_south <- crop(twi, south_shed) 


twi_south_mask <- mask(twi, south_shed)


mapview(twi_south_mask)


global(twi_south_mask, fun = 'mean', na.rm = T)
```

# Assignment

## Q1 Generate your own watershed

Using the above code as a starter, generate your own watershed. Try to only pick a watershed that is not too large in size (\~ \< 1000km2). Remember you may need to mess around with the Z level for get_elev_raster().

```{r}


points <- tibble( site = c("West_Branch", 
                           "North_Branch"), 
                  lat_long = c("40.6724777,-105.8620178", 
                               "40.74148,-105.8924542"))%>%
  separate(lat_long, into = c("lat", "long"), sep = ",")%>%
  sf::st_as_sf(coords = c("long","lat"), crs = 4326)%>%
  st_transform(2163)
  

preview_dem <- function(site_name, lat_long){
  
  # Create a directory for the site
  site_dir <- paste0("data/sjs/", site_name)
  
  #create folder for tifs
  if(!file.exists(site_dir)){
  dir.create(site_dir)
  }
  
  #Create a pourpoint
  points <- tibble( site = site_name, 
                    lat_long = lat_long)%>%
    separate(lat_long, into = c("lat", "long"), sep = ",")%>%
    sf::st_as_sf(coords = c("long","lat"), crs = 4326)%>%
    st_transform(2163)%>%
  st_write(., paste0(site_dir, "/location.shp"), append = F)
  
  #prompt user for zoom level
  z <- readline(prompt = "What zoom level would you like to use? (1-15)  ")
  # download dem with elevatr and zoom level
  dem <- points%>%
    get_elev_raster(., z = as.numeric(z)) %>%
    rast(.)
  #Confirm if zoom level is appropriate
 mapview(dem)+ mapview(points)

}

# this function will show you your point and the dem you are about to download based on the zoom level indicated by user
#preview_dem("West_Branch", "40.6724777,-105.8620178")


```

# Big function

```{r}


create_watersheds <- function(site_name, lat_long, zoom_level = 11){
  
    
  # Create a directory for the site
  site_dir <- paste0("data/sjs/", site_name)
  #create folder for tifs
  if(!file.exists(site_dir)){
  dir.create(site_dir)
  }

  #read in pour point
 points <- tibble( site = site_name, 
                    lat_long = lat_long)%>%
    separate(lat_long, into = c("lat", "long"), sep = ",")%>%
    sf::st_as_sf(coords = c("long","lat"), crs = 4326)%>%
    st_transform(2163)%>%
  st_write(., paste0(site_dir, "/location.shp"), append = F)

  # download dem with elevatr and zoom level
  dem <- points%>%
    get_elev_raster(., z = as.numeric(zoom_level)) %>%
    rast(.)
  #create DEM
  dem_dir <- paste0(site_dir, '/elev.tif')
  writeRaster(dem, dem_dir, overwrite = T)
  
  
  # Breach depressions (force water to move through ponds/lakes/etc...)
wbt_breach_depressions_least_cost(
  dem = dem_dir,
  output = paste0(site_dir, "/breached.tif"),
  dist = 9,
  fill = TRUE)


# Fill any remaining depressions
wbt_fill_depressions_wang_and_liu(
  dem = paste0(site_dir, "/breached.tif"),
  output = paste0(site_dir, "/breachfill.tif")
)

#Get flow accumulation
wbt_d8_flow_accumulation(input = paste0(site_dir, "/breachfill.tif"),
                         output = paste0(site_dir, "/d8fa.tif"))
#Get flow direction
wbt_d8_pointer(dem = paste0(site_dir, "/breachfill.tif"),
               output = paste0(site_dir, "/d8point.tif"))

#Extract streams at 

wbt_extract_streams(flow_accum = paste0(site_dir, "/d8fa.tif"),
                    output = paste0(site_dir, "/raster_streams.tif"),
                    threshold = 1000)

#Snap our watershed flowliens to this point. 
wbt_jenson_snap_pour_points(pour_pts = paste0(site_dir, "/location.shp"),
                            streams = paste0(site_dir, "/raster_streams.tif"),
                            output = paste0(site_dir, "/snapped_location.shp"),
                            snap_dist = 200)

wbt_watershed(d8_pntr = paste0(site_dir, "/d8point.tif"),
              pour_pts = paste0(site_dir, "/snapped_location.shp"),
              output = paste0(site_dir, "/watershed.tif"))

watershed <- rast(paste0(site_dir, "/watershed.tif"))%>%
  trim()

mapview(watershed)+ mapview(points)
}

#create_watersheds("West_Branch","40.6724777,-105.8620178",  11)

watersheds <- tibble(site_name = c("West_Branch", 
                           "North_Branch", 
                           "Glacier_Gorge" ), 
                      lat_long = c("40.6724777,-105.8620178", 
                               "40.74148,-105.8924542", 
                               "40.310325,-105.6373891"), 
                     zoom_level = c(10,10, 12))


 pmap(watersheds, create_watersheds)


```



## Generate, visualize, and extract two additional terrain metrics (not TWI)

```{r}


 calc_mean_ruggedness <- function(site_name){
   
  watershed <- rast(paste0("data/sjs/",site_name,"/watershed.tif"))
   
  wbt_ruggedness_index(dem = paste0("data/sjs/",site_name,"/elev.tif"), 
                               output = paste0("data/sjs/",site_name,"/ruggedness.tif"))
  ruggedness_ws <- rast(paste0("data/sjs/",site_name,"/ruggedness.tif"))%>%
  crop(.,watershed)%>%
  mask(.,watershed)
  
  writeRaster(ruggedness_ws, paste0("data/sjs/",site_name,"/ruggedness_ws.tif"), overwrite = T)
  
 global(ruggedness_ws, fun = 'mean', na.rm = T)
 
 }

 calc_mean_aspect <- function(site_name){
   
  watershed <- rast(paste0("data/sjs/",site_name,"/watershed.tif"))
   
  wbt_aspect(dem = paste0("data/sjs/",site_name,"/elev.tif"), 
                               output = paste0("data/sjs/",site_name,"/aspect.tif"))
  
  aspect_ws <- rast(paste0("data/sjs/",site_name,"/aspect.tif"))%>%
  crop(.,watershed)%>%
  mask(.,watershed)
  
  writeRaster(aspect_ws, paste0("data/sjs/",site_name,"/aspect_ws.tif"), overwrite = T)
  
 global(aspect_ws, fun = 'mean', na.rm = T)
 
 }

calc_mean_slope <- function(site_name){
   
  watershed <- rast(paste0("data/sjs/",site_name,"/watershed.tif"))
   
  wbt_slope(dem = paste0("data/sjs/",site_name,"/elev.tif"), 
                               output = paste0("data/sjs/",site_name,"/slope.tif"))
  slope_ws <- rast(paste0("data/sjs/",site_name,"/slope.tif"))%>%
  crop(.,watershed)%>%
  mask(.,watershed)
  
  writeRaster(slope_ws, paste0("data/sjs/",site_name,"/slope_ws.tif"), overwrite = T)
  
  global(slope_ws, fun = 'mean', na.rm = T)
 
 }

watershed_calcs <- watersheds %>%
  mutate(ruggedness = map(site_name, calc_mean_ruggedness), 
         slope = map(site_name, calc_mean_slope), 
         aspect = map(site_name, calc_mean_aspect))%>%
  unnest(cols = c(ruggedness))%>%
  rename(mean_ruggedness = mean)%>%
  unnest(cols = c(slope))%>%
  rename(mean_slope = mean)%>%
   unnest(cols = c(aspect))%>%
  rename(mean_aspect = mean)

view_ruggedness <- function(site_name){
  ruggedness <- rast(paste0("data/sjs/",site_name,"/ruggedness_ws.tif"))
  mapview(ruggedness)
}
view_slope <- function(site_name){
  slope <- rast(paste0("data/sjs/",site_name,"/slope_ws.tif"))
  mapview(slope)
}
view_aspect <- function(site_name){
  aspect <- rast(paste0("data/sjs/",site_name,"/aspect_ws.tif"))
  mapview(aspect)
}
map( watersheds$site_name, view_ruggedness)
map( watersheds$site_name, view_slope)
map( watersheds$site_name, view_aspect)



```

## Aspect slope wind rose plot

```{r}
#using the slope and aspect rasters, create a plot with 360 degrees on the x-axis and the mean slope on the y-axis.
#The plot should have a line for each site, and the lines should be colored by site.

slope_aspect_plot <- function(site_name){
  
  watershed <- rast(paste0("data/sjs/",site_name,"/watershed.tif"))
  aspect <- rast(paste0("data/sjs/",site_name,"/aspect_ws.tif"))
  slope <- rast(paste0("data/sjs/",site_name,"/slope_ws.tif"))
  
  aspect <- as.matrix(aspect)
  slope <- as.matrix(slope)
  
  aspect <- aspect[!is.na(aspect)]
  slope <- slope[!is.na(slope)]
  
  slope_aspect <- tibble(aspect = aspect, slope = slope)%>%
    mutate(aspect = round(aspect, digits = 0), 
           aspect = case_when(aspect == 0 ~ 360, 
                              TRUE ~ aspect))%>%
    filter(aspect != -1)
  
  slope_aspect <- slope_aspect%>%
    group_by(aspect)%>%
    summarize(mean_slope = mean(slope))
  
  ggplot(slope_aspect, aes(x = aspect, y = mean_slope))+
  # make this a polar plot with coord_polar with 0b beint at the top and 180 at the bottom
    geom_line()+
    coord_polar(start = 0, theta = "x")+
    theme_minimal()+
    labs(title = site_name, x = "Aspect (degrees)", y = "Mean Slope")+
    # add a label at 180 degrees called "S"
    geom_text(x = 180, y = 0, label = "N", size = 5)+
    # add a label at 0 degrees called "N"
    geom_text(x = 0, y = 0, label = "S", size = 5)+
    # add a label at 90 degrees called "E"
    geom_text(x = 90, y = 0, label = "W", size = 5)+
    # add a label at 270 degrees called "W"
    geom_text(x = 270, y = 0, label = "E", size = 5)
}
GG_sa <- slope_aspect_plot("Glacier_Gorge")+
   theme(axis.title.x = element_blank())
WB_sa <- slope_aspect_plot("West_Branch")+
  #remove X axis label
  theme(axis.title.x = element_blank())
NB_sa <- slope_aspect_plot("North_Branch")


ggpubr::ggarrange(GG_sa, WB_sa, NB_sa, nrow = 3)

```


