---
title: "Terrain Analyses in R"
author: "Matthew Ross"
date: "2024-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(terra)
library(whitebox)
library(elevatr)
library(mapview)
library(tidyverse)
library(tmap)


# Run first time you install whitebox
#whitebox::install_whitebox()
```


# Goals


## Terrain Analysis Toolkit

The goal of this lab is to familiarize you with a very powerful set of tools
that enable you to do terrain analyses. These are primarily the packages `elevatr` for
downloading DEM data, `whitebox` for conducting watershed and terrain analyses.
These packages also rely on other geospatial packages (`sf`, `terra`) and visualizing
the data relies on `mapview` and `tmap`. 


## Terrain Analysis Ideas

So the above tools can help you analyze a digital elevation model, but why would
you want to do this? Well, terrain analysis can do many things in water resource
planning and analysis. For example, if we wanted to find parts of the landscape
where we think wetlands could exist, we might want to use [Topographic Wetness 
Index](https://www.whiteboxgeo.com/manual/wbt_book/available_tools/geomorphometric_analysis.html#wetnessindex) 
to identify these areas where the topography lends itself to wet landscapes, which
are areas of relatively high flow accumulation coupled with a bowl-like curvature.

To learn these tools and ideas we will work through a watershed delineation example
in the South Fork of the Poudre River, and then you will conduct your own analysis. 


# South Fork Example

## Add a watershed outlet point

```{r}

#Create a point near the stream using lat, long. 
sheds <- tibble(site = c('South Fork'),
                long = c(40.5475185),
                lat = c(-105.6091385)) %>%
  st_as_sf(., coords = c('lat','long'), crs = 4263) %>%
  st_transform(2163)

#Visualize the points
mapview(sheds)

#Little code snippet to make the data folder if you don't have it. 
if(!file.exists('data')){
  dir.create('data')
}

st_write(sheds, 'data/pourpoints.shp', append = F)
```


## Get elevation data from around that area. 

```{r}

# Use elevatr to download data and convert to terra object
elev <- get_elev_raster(sheds, z = 10) %>%
  rast(.)

# Various ways to plot

mapview(elev) + 
  mapview(sheds)
```

### Save a local DEM Version

```{r}
#Save local version for whitebox to use
writeRaster(elev, 'data/elev.tif', overwrite = T)
```



## Get hillshade for display purposes

```{r}

#Get a shillshade map from wbt
wbt_hillshade(dem = "data/elev.tif",
              output = "data/hillshade.tif",
              azimuth = 115)

#Read in the hillshade
hillshade <- rast('data/hillshade.tif')

#Visualize
tm_shape(hillshade)+
  tm_raster(style = "cont",
            palette = "-Greys", 
            legend.show = FALSE)+
  tm_scale_bar()
```


## Hydrologically condition DEM

JP Gannon does a great job on his [hydroinformatics website](https://vt-hydroinformatics.github.io/Quarto_Book/14-Geospatial-Raster-Hydro.html) explaining why we need to hydrologically condition a DEM before we can conduct
watershed analyses. Basically, DEMs aren't perfect and they can have artifacts 
in them that make it so that water doesn't properly flow downhill. To force water
to move the way we think it should through the landscape we can "condition" these
DEMs using the functions below. 

```{r}

# Breach depressions (force water to move through ponds/lakes/etc...)
wbt_breach_depressions_least_cost(
  dem = "data/elev.tif",
  output = "data/breached.tif",
  dist = 9,
  fill = TRUE)


# Fill any remaining depressions
wbt_fill_depressions_wang_and_liu(
  dem = "data/breached.tif",
  output = "data/breachfill.tif"
)
```

## Flow Accumulation 

Flow accumulation is the key function we use to estimate how much watershed area
is draining to a specific point anywhere in the river network. Here we use the D8
flow algorithm, which makes it so 100% of the water is routed to nearby cells. 
ESRI has a nice article visualizing what this algorithm is [doing](https://pro.arcgis.com/en/pro-app/latest/tool-reference/raster-analysis/flow-direction.htm)

```{r}

#Get flow accumulation
wbt_d8_flow_accumulation(input = "data/breachfill.tif",
                         output = "data/d8fa.tif")
#Get flow direction
wbt_d8_pointer(dem ='data/breachfill.tif',
               output = 'data/d8point.tif')

# read in the data
fa <- rast('data/d8fa.tif') %>%
  log10(.)

#visualize
mapview(fa) + 
  mapview(sheds)
```


## Extract streams with arbitrary 300 cell threshold 

In order to make sure that our watershed outlet point, which we arbitrarily and
manually extracted lats and longs for, we need to first extract a stream network.
We are picking a 300-cell flow accumulation threshold for our 'stream initiation.' 
Because our starting raster is a ~58X58m cell size 300 cells roughly equals 1 km2.
This means, we are saying that once a stream has more than 1 km2 of watershed area draining
to that point in the stream, we think that it will have a functioning stream channel.
You can conduct entire studies to get this threshold right, so this is arbitrary here. 

```{r}

#Extract streams at 1km2 
wbt_extract_streams(flow_accum = "data/d8fa.tif",
                    output = "data/raster_streams.tif",
                    threshold = 300)

#Snap our watershed flowliens to this point. 
wbt_jenson_snap_pour_points(pour_pts = "data/pourpoints.shp",
                            streams = "data/raster_streams.tif",
                            output = "data/snappedpp.shp",
                            snap_dist = 200)

# Read in and check the snap. 
snap_pour <- st_read('data/snappedpp.shp')
mapview(fa) + 
  mapview(snap_pour)

```


## Delineate watershed


Finally after all that! We can delineate our watershed


```{r}
wbt_watershed(d8_pntr = "data/d8point.tif",
              pour_pts = "data/snappedpp.shp",
              output = "data/south_fork.tif")

south_shed <- rast('data/south_fork.tif')

mapview(south_shed) + 
mapview(sheds)

```

## Extract a terrain metric about this watershed

### Generate terrain metric

Delineating a watershed allows us to now get watershed terrain metrics for that 
watershed. Whitebox tools has dozens of terrain tools you can use
[See More in the Geomorph Section](https://www.whiteboxgeo.com/manual/wbt_book/intro.html).
Here we will generate and extract one critical variable. Topographic Wetness,
which indicates areas in a watershed (or a watershed average) wetness. High values
indicate areas that can be marshy/wetlandy and generally wet. Low values (like ridge tops) will
be drier parts of the landscape. 

```{r}
#First we need slope
wbt_slope(dem = 'data/elev.tif',
          output = 'data/slope.tif',
          units = 'degrees')

# Now we can generate twi
wbt_wetness_index(sca = 'data/d8fa.tif',
                  slope = 'data/slope.tif',
                  output = 'data/twi.tif')

twi <- rast('data/twi.tif')

mapview(twi)


```

### Crop to watershed and extract average value


```{r}
# Crop TWI to south_shed area

twi_south <- crop(twi, south_shed) 


twi_south_mask <- mask(twi, south_shed)


mapview(twi_south_mask)


global(twi_south_mask, fun = 'mean', na.rm = T)
```



# Assignment


## Q1 Generate your own watershed

Using the above code as a starter, generate your own watershed. Try to only
pick a watershed that is not too large in size (~ < 1000km2). Remember
you may need to mess around with the Z level for get_elev_raster(). 

Teton water shed

```{r}

#Create a point near the stream using lat, long. 
sheds_2 <- tibble(site = c('Teton Leaigh watershed'),
                long = c(43.7624919),
                lat = c(-110.7776054)) %>%
  st_as_sf(., coords = c('lat','long'), crs = 4263) %>%
  st_transform(2163)

#Visualize the points
mapview(sheds_2)

#Little code snippet to make the data folder if you don't have it. 
if(!file.exists('data')){
  dir.create('data')
}

st_write(sheds_2, 'data/pourpoints_teton.shp', append = F)
```

```{r}

# Use elevatr to download data and convert to terra object
elev_2 <- get_elev_raster(sheds_2, z = 11) %>%
  rast(.)

# Various ways to plot

mapview(elev_2) + 
  mapview(sheds_2)
```

```{r}
#Save local version for whitebox to use
writeRaster(elev_2, 'data/elev_teton.tif', overwrite = T)
```



## Get hillshade for display purposes

```{r}

#Get a shillshade map from wbt
wbt_hillshade(dem = "data/elev_teton.tif",
              output = "data/hillshade_teton.tif",
              azimuth = 115)

#Read in the hillshade
hillshade_teton <- rast('data/hillshade_teton.tif')

#Visualize
tm_shape(hillshade_teton)+
  tm_raster(style = "cont",
            palette = "-Greys", 
            legend.show = FALSE)+
  tm_scale_bar()
```

## Generate, visualize, and extract two additional terrain metrics (not TWI)

```{r}

```

```{r}

# Breach depressions (force water to move through ponds/lakes/etc...)
wbt_breach_depressions_least_cost(
  dem = "data/elev_teton.tif",
  output = "data/breached_teton.tif",
  dist = 9,
  fill = TRUE)


# Fill any remaining depressions
wbt_fill_depressions_wang_and_liu(
  dem = "data/breached_teton.tif",
  output = "data/breachfill_teton.tif"
)
```



```{r}

#Get flow accumulation
wbt_d8_flow_accumulation(input = "data/breachfill_teton.tif",
                         output = "data/d8fa_teton.tif")
#Get flow direction
wbt_d8_pointer(dem ='data/breachfill_teton.tif',
               output = 'data/d8point_teton.tif')

# read in the data
fa_teton <- rast('data/d8fa_teton.tif') %>%
  log10(.)

#visualize
mapview(fa_teton) + 
  mapview(sheds_2)

library(terra)
elev_teton <- rast('data/elev_teton.tif')
cell_size <- res(elev_teton)
print(cell_size)
```



```{r}
#Save local version for whitebox to use
writeRaster(elev_teton, 'data/elevteton.tif', overwrite = T)

```


```{r}

#Extract streams at 1km2 
wbt_extract_streams(flow_accum = "data/d8fa_teton.tif",
                    output = "data/raster_streams_teton.tif",
                    threshold = 700)

```

```{r}


#Snap our watershed flowliens to this point. 
wbt_jenson_snap_pour_points(pour_pts = "data/pourpoints_teton.shp",
                            streams = "data/raster_streams_teton.tif",
                            output = "data/snappedpp_teton.shp",
                            snap_dist = 200)


```



## Delineate watershed



```{r}
wbt_watershed(d8_pntr = "data/d8point_teton.tif",
              pour_pts = "data/snappedpp_teton.shp",
              output = "data/Teton.tif")

Teton_shed <- rast('data/Teton.tif')

mapview(Teton_shed) + 
mapview(sheds_2)

```

```{r}
# Generating the Aspect of the terrain
wbt_aspect(
    dem = "data/elev_teton.tif",
    output = "data/asp_teton.tif",
    zfactor = 1  
)


# Read in the aspect data
aspect_teton<- rast("data/asp_teton.tif")

mapview(aspect_teton)


# Crop aspect to shed_2 area
cropped_aspect_teton <- terra::crop(aspect_teton, Teton_shed) 

# mask aspect to remove external data outside of watershed area
aspect_teton_mask <- terra::mask(aspect_teton, Teton_shed)

mapview(aspect_teton_mask) + mapview(Teton_shed)

#extract mean aspect value from watershed
global(aspect_teton_mask, fun = 'mean', na.rm = T)

```


```{r}
#classifying landform using Geomorphon tool
wbt_geomorphons(
    dem = "data/elev_teton.tif",
    output = "data/geo_teton.tif",
    search=50, 
    threshold=0.0, 
    fdist=0, 
    skip=0,
    forms="True",
    residuals="False"
    
)


# Read in the geomorphone data
geomorphons<- rast("data/geo_teton.tif")

# Crop geomorphone to teton area
cropped_geo_teton <- terra::crop(geomorphons, Teton_shed) 

# mask geomorhone to remove external data outside of watershed area
geo_teton_mask <- terra::mask(geomorphons, Teton_shed)


#Classify the landfrom in distinct different colors and make the Twton shed area transparent. 

distinct_colors <- c("#e6194B", "#3cb44b", "#ffe119", "#4363d8", "#f58231", 
                     "#911eb4", "#46f0f0", "#f032e6", "#bcf60c", "#fabebe",
                     "#008080", "#e6beff", "#9a6324", "#fffac8", "#800000",
                     "#aaffc3", "#808000", "#ffd8b1", "#000075", "#808080", 
                     "#ffffff", "#000000")

mapview(geo_teton_mask, col.regions = distinct_colors) + mapview(Teton_shed, alpha.regions = 0.1,legend = FALSE)

#here Geomorphone tool from white box tool is used for landform classification based on a DEM. I was not able to edit the label in the scale bar butthe ledgend for the scale bar is as followed: 
#Value	Landform Type
#1	Flat
#2	Peak (summit)
#3	Ridge
#4	Shoulder
#5	Spur (convex)
#6	Slope
#7	Hollow (concave)
#8	Footslope
#9	Valley
#10	Pit (depression)

```
